Towards a Mechanized Proof of Noninterference

### 2013-12-18 paper/paper.tex
Corrections, more writing, a todo macro, and some explanation of related work and our experience with mechanization.

\newcommand{\langName}[0]{TinyML\textsuperscript{2}}
\newcommand{\origLang}[0]{CoreML\textsuperscript{2}}
\newcommand{\typeRule}[3]{#1 \vdash #2 \colon #3}
\newcommand{\guards}[0]{\lhd}
\newcommand{\lift}[1]{\lfloor #1 \rfloor}

@@

\begin{figure}[h!]
  \begin{align}
    v & ::= x \: | \: k \: | \: \textsf{fix} f . \lambda x . e \tag{values} \: |
    \: <v | v> \\
    a & ::= v \tag{answers} \: | \: <a | a> \\
    e & ::= a \: | \: v v \: | \: \textsf{if } v \: e \: e \: | \: \textsf{let
    } x := v \textsf{ in } e \tag{expressions} \: | \: <e | e>
  \end{align}
  \caption{Syntax of \langName}
  \label{fig:syntax}
\end{figure}

@@

\begin{lemma}[Completeness]
  \label{lem:completeness}
  Assume $\lift{e}_i \to^* a_i$ for all $i \in \{1, 2\}$.  Then, there exists an
  answer $a$ such that $e \to^* a$.
\end{lemma}
\begin{proof}
  \textbf{TODO}
\end{proof}

@@

@@

\begin{lemma}
  Let $H$ be an upward-closed subset of the label lattice $\mathcal{L}$.  Let $l
  \not\in H$.  If, using $H$ for the set of high labels, $\typeRule{
  }{e}{\textsf{int}^l}$ and $e \to^* v$ then $\lift{v}_1 = \lift{v}_2$.
\end{lemma}
\begin{proof}
  \textbf{TODO}
\end{proof}

@@

\begin{thm}[Noninterference]
  Choose $l, h \in \mathcal{L}$ such that $h \not\leq l$.  Let $h \guards t$.
  Assume $\typeRule{(x \mapsto t)}{e}{\textsf{int}^l}$, where $e$is a \langName
  expression.  If, for all $i \in \{1, 2\}$, $\typeRule{ }{v_i}{t}$ and $e[x
  \Leftarrow v_i] \to^* v_i'$ hold, then $v_1' = v_2'$.
\end{thm}
\begin{proof}
  \textbf{TODO}
\end{proof}

@@

Talk about what did/should have been mechanized.  Talk about troubles and/or
interesting bits here.  Might point out here that we're fairly certain that we
found an inaccuracy in the original paper.

@@

Discuss various references related to our project.

Discuss:

\begin{itemize}
  \item Proof carrying code
  \item Proof assistants
  \item Verified languages and compilers
  \item Information flow languages
\end{itemize}

@@

Talk about stuff we still have to do for this to be a complete project.  Talk
about expected problems and required techniques.  Talk about what we could do
once this project is completed and we could use it for other projects.

@@

Summarize again what we covered, where we ended up, and where we'd like to go
next.

>>>>

\newcommand{\langName}[0]{TinyML\textsuperscript{2}}
\newcommand{\origLang}[0]{CoreML\textsuperscript{2}}
\newcommand{\typeRule}[3]{#1 \vdash #2 \colon #3}
\newcommand{\guards}[0]{\lhd}
\newcommand{\lift}[1]{\lfloor #1 \rfloor}
\newcommand{\todo}[1]{\textbf{TODO: #1}}

@@

\begin{figure}[h!]
  \begin{align}
    v & ::= x \: | \: k \: | \: \textsf{fix} f . \lambda x . e \tag*{ } \\
    v' & ::= v \: | \: <v | v> \tag{values}\\
    a & ::= v \tag*{ } \\
    a' & ::= a \: | \: <a | a> \tag{answers} \\
    e & ::= a \: | \: v v \: | \: \textsf{if } v \: e \: e \: | \: \textsf{let
    } x := v \textsf{ in } e \tag*{ } \\
    e' & ::= e \: | \: <e | e> \tag{expressions}
  \end{align}
  \caption{Syntax of \langName.  The prime variants are the full class of
  expressions, values, and answers.  This construction avoids allowing nested
  branching values.}
  \label{fig:syntax}
\end{figure}

@@

\begin{lemma}[Completeness]
  \label{lem:completeness}
  Assume $\lift{e}_i \to^* a_i$ for all $i \in \{1, 2\}$.  Then, there exists an
  answer $a$ such that $e \to^* a$.
\end{lemma}
\begin{proof}
  First, we establish that if $\lift{e}_i$ reduces to an answer in a finite
  number of steps, then $e$ does not have an infinite reduction sequence.  The
  only way this could occur is if $e$ was infinitely reduced over lift rules,
  because otherwise, since the only reductions on $e$ that correspond to 0
  reductions on either projection are lift reduction rules.  But as we can see
  from inspecting the lift reduction rules, we cannot continuously apply these
  rules exclusively.  Thus, we can conclude that $e$ does not have an infinite
  series of reduction steps that can be applied to it.

  As a result, we know that $e$ reduces to an answer, because otherwise it would
  be stuck and this would mean that one of the projections is also stuck by
  Lemma~\ref{lem:bracket_stuck}, a contradicton.  Thus, we know that if both
  projections step to an answer, so must the original expression.
\end{proof}

@@

The following lemma asserts that if a value is well typed, the expression
containing just that value is also well typed.

\begin{lemma}[Value]
  \label{lem:value}
  $\typeRule{pc}{v}{t}$ implies $\typeRule{ }{v}{t}$
\end{lemma}
\begin{proof}
  By induction on the proof of $\typeRule{pc}{v}{t}$.  The cases are either that
  we have \textsc{E-Value}, where we're done, \textsc{E-Sub}, in which case we
  apply the inductive hypothesis and the rule \textsc{V-Sub}, or we have
  \textsc{E-Bracket}, which is proved by the inductive hypothesis and
  \textsc{V-Bracket}.
\end{proof}

We also prove a subject reduction lemma, allowing us to assert that if an
expression is well typed and there is a reduction that can apply to it, then the
reduced form is also well typed, with the same type.

\begin{lemma}[Subject Reduction]
  \label{lem:subj_red}
  Let $e \to e'$.  Assume $\typeRule{pc}{e}{t}$ and $pc \in H$ for all $i \in
  \{1, 2\}$.  Then $\typeRule{pc}{e'}{t}$.
\end{lemma}
\begin{proof}
  \todo{}
\end{proof}

@@

\begin{lemma}
  \label{lem:final_support}
  Let $H$ be an upward-closed subset of the label lattice $\mathcal{L}$.  Let $l
  \not\in H$.  If, using $H$ for the set of high labels, $\typeRule{
  }{e}{\textsf{int}^l}$ and $e \to^* v$ then $\lift{v}_1 = \lift{v}_2$.
\end{lemma}
\begin{proof}
  By Lemma~\ref{lem:subj_red} and \ref{lem:value}, we know that $\typeRule{
  }{v}{\textsf{int}^l}$ holds.  Since the type is $\textsf{int}^l$, we know that
  $v$ is either $k$ or $< k_1 | k_2 >$.  However, we know that if it were the
  latter, then by \textsc{V-Bracket} this would imply there is an $l' \leq l$
  such that $l' \in H$, a contradiction.  Thus, $v$ must be of the form $k$
  which is the same under either projection.
\end{proof}

@@

\begin{thm}[Noninterference]
  Choose $l, h \in \mathcal{L}$ such that $h \not\leq l$.  Let $h \guards t$.
  Assume $\typeRule{(x \mapsto t)}{e}{\textsf{int}^l}$, where $e$is a \langName
  expression.  If, for all $i \in \{1, 2\}$, $\typeRule{ }{v_i}{t}$ and $e[x
  \Leftarrow v_i] \to^* v_i'$ hold, then $v_1' = v_2'$.
\end{thm}
\begin{proof}
  Let $H$ be the upward closure of $\{h\}$ and $v = < v_1 | v_2 >$, since other
  $v$ would be immediately uninteresting.  By \textsc{V-Bracket}, we know that
  $\typeRule{ }{v}{t}$ holds given the assumptions.
  By Lemma~\ref{lem:substitution}, we have $\typeRule{ }{e[x \Leftarrow
  v]}{\textsf{int}^l}$.  By Lemma~\ref{lem:final_support}, we know that
  $\lift{e[x \Leftarrow v]}_i = e[x \Leftarrow \lift{v}_i]$, which under the
  current assumptions reduces to $v_i$.  By Lemma~\ref{lem:completeness} we know
  that there is an answer $a$ such that $e[x \Leftarrow v] ->^* a$.  By
  Lemma~\ref{lem:soundness}, $\lift{a}_i = v_i'$ for all $i$ which means that
  $a$ is a value.  Then, we know by Lemma~\ref{lem:final_support} that $v_1' =
  v_2'$.
\end{proof}

@@

As mentioned above, we unfortunately were unable to complete our noninterference
proof in Coq by the time of this report.  Given more time, we are confident that
this could be completed however.  So far, we have proven the first few lemmas
described in Section~\ref{sec:proof}.  This was much more difficult than the
short natural language proofs would suggest, due to a combination of the
stricter format required by Coq and the number of items that are elided in the
proofs because they were considered trivial or obvious.

We found that the process of mechanizing the lemmas that we proved was a useful
exercise in checking our understanding of our language, statements, and the work
of Simonet and Pottier which we based our approach off of.  We found ourselves
often surprised by the number of items that were elided by the proofs in the
paper.  For example, the proof of Lemma~\ref{lem:soundness} is one line of
natural languae but turned out to be 40 lines of Coq code, not including all of
the supporting definitions and rules.

Our proof of Lemma~\ref{lem:soundness} actually turned out to be instructive
because it forced us to recognize that the original formulation in Simonet and
Pottier's paper on \origLang~\cite{InfoFlowML} wasn't quite correct.
Specifically, it is entirely possible that a step in the operational semantics
does not translate to a step in the operational semantics under either
projection, as in the case of the \ref{eq:lift-app} rule, where $\lift{e}_i$ and
$\lift{e'}_i$ are identical.  Luckily, we are fairly confident, given that we
were still able to prove Lemma~\ref{lem:bracket_stuck} despite this change, that
this is not a critical error in the original paper.

\todo{More.}

@@

This project is related to a variety of previous work, specifically in the realm
of security and verified proofs and programs.  We are following in the footsteps
of verified languages and compilers, which focus on writing programs and
language specs that are verified to have the properties the authors want by
providing the proofs of these properties on the
definitions~\cite{strecker2002formal,leinenbach2005towards,dave2003compiler,leroy2009formally,leroy2006formal,leroy2009formal,leroy2012compcert,chlipala2010verified,berghofer2004extracting}.
This has been a popular area, which is primarily focused on more mainstream
languages and compilers (such as C and Java).

Another related avenue of research has been the work on proof carrying code,
which involves a compiler producing a proof that the program satisfies some
property along with the compiled program, allowing a smaller trusted computing
base for security sensitive computing
scenarios~\cite{necula2002proof,necula1998design}.  Our goal was to make a step
towards producing a verifying compiler for a language with information flow
types like Jif or FlowCaml, which use a type system very similar to the one
presented in this paper for ensuring noninterference of
programs~\cite{InfoFlowML,myers1999jflow}.

@@

\todo{Talk about stuff we still have to do for this to be a complete project.
Talk about expected problems and required techniques.  Talk about what we could
do once this project is completed and we could use it for other projects.}

@@

\todo{Summarize again what we covered, where we ended up, and where we'd like to
go next.}
